Cognate beginnings to bilingual lexical acquisition

### 2023-06-15 manuscript/manuscript.qmd
Dockerise repository

abstract: |
    Recent work on bilingual language acquisition points to cognate words being acquired earlier in age than non-cognates. Cross-language activation has been proposed as a candidate mechanism behind this facilitation effect, but the

       but the specific dynamics of lexical Building on accumulator models of language acquisition, we present an account of word acquisition in which the co-activation of cognate translation equivalents increases the rate at which of bilingual children encounter learning instances for words in both languages. We tested these predictions against vocabulary data from 366 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the participant. We found a cognateness facilitation effect only in low-frequency words or words from the language of lower exposure. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to both experimental data [@jusczyk1995infants; @tincoff1999beginnings; @bergelson2012months; @bergelson2015early] and parental vocabulary reports [e.g., @fenson2007macarthurbates; @samuelson2021precision]. During the last half of the second year of life, new word-forms are acquired at an increasingly fast rate [@goldfield1990early; @fenson1994variability; @mayor2011statistical; @bloom2002how; @bergelson2020comprehension]. Such early stages of lexical acquisition are characterised by substantial heterogeneity across different populations of language learners, reflected for instance on (the variability of) quantitative indices of lexical development, like vocabulary size [i.e., vocabulary size, @fenson1994variability; @frank2021variability]. Despite this variability, trajectories of early vocabulary growth seem quite stable across languages. @tardif2008baby collected data about the first ten words acquired by 10-- to 16--month-old infants living in the United States, Hong Kong, and Beijing. Since birth, these infants had been learning English, Cantonese, and Mandarin, respectively. The authors found a common pattern across the three groups: their first ten words referred to roughly the same concepts, namely relatives/caregivers (*daddy*, *mommy*), social routines (*bye*, *uh-oh*) or animal sounds (*woof-woof*). These results were later extended by @frank2021variability to a more diverse set of languages, and found that such cross-language commonalities are stronger at earlier stages of lexical acquisition.

Most research on early word acquisition relies extensively on data from monolingual children, and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous research on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and also to both groups knowing a similar amount of words when the bilinguals' two languages are taken into account. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together, which substantiates the importance of collecting data on both languages when assessing bilinguals' communicative development. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. These studies rely on bilingual samples whose languages are relatively distant in typology: English is a Germanic language, while Spanish is a Romance language, and both languages differ substantially at the grammatical and the phonological level. Little is known about vocabulary acquisition in bilinguals learning a pair of typologically closer languages (e.g., when both are Romance languages).

One way typological distance may impact bilingual trajectories of language acquisition is the presence (or absence) of *cognates* in their linguistic input. Quite often, bilingual children encounter more than one label for the same referent: one in each language. For instance, an English-Spanish bilingual might hear both *dog* and *perro* in the presence of a dog We will refer to such cross-language synonyms as *translation equivalents*.^[Throughout the present study, we make the assumption that bilinguals form a common semantic representations for each translation equivalent (e.g., the concept 'dog' is shared for the English and Spanish words *dog* and *perro*). The referential contexts in which bilingual children hear (and learn) both labels may not be identical. For instance, the English word *finger* refers to eight body parts (excluding thumbs and toes), whereas its Spanish translation *dedos* refers to twenty body parts (fingers, thumbs, and toes). These discrepancies are reflected, for instance, in adults' performance in lexical decision tasks [@chaouch-orozco2023are]. The fact that some translation pairs share more semantic features than others does not, however, inform whether bilinguals form common or separate semantic representations for translation equivalents. What English speakers understand for 'fingers' might not be identical to what Spanish speakers understand for 'dedos'. But when an English speaker learns Spanish, it seems implausible that they would form a separate semantic representation for what 'dedo' conveys at the conceptual level, compared to the possibility that they associate the new label to the semantic representation they already had formed for 'fingers'. Therefore, the lack of complete semantic alignment is not necessarily an argument against the existence of common semantic representations for translation equivalents. For simplicity, and in line with previous (formal and verbal) models of bilingual lexical processing [e.g., @shook2013bilingual; @dijkstra2019multilink; @dijkstra1998bia; @kroll2010revised], we assume common semantic representations for translation pairs.]. Sometimes, the word-forms involved in a translation equivalent may be phonologically similar. Such form-similar translation equivalents are referred to as *cognates*. While the previously mentioned translation pair *dog* and *perro* are non-cognates (both word-forms hardly share any phonemes), the translation pair *cat* and *gato* can be considered cognates. The amount of cognates shared by a pair of languages is dependent on their typological distance. When both languages are typologically close, like Spanish and Catalan (both Romance languages), they are more likely to share a large amount of cognates (i.e., form-similar translation equivalents) than two typologically more distant languages, like English and Spanish (Romance and Germanic languages, respectively). Therefore, bilingual children learning two typologically close languages may be exposed to a larger amount of cognates than those learning two typologically more distant languages. For instance, in the presence of a door, a Catalan-Spanish bilingual might be exposed to the words *porta* and *puerta* (cognate translation equivalent with the common meaning 'door'), whereas an English-Spanish bilingual might hear *door* and *puerta* (non-cognate translation equivalent).

It could be the case that mapping two phonologically similar labels (cognates like *porta*-*puerta*) onto the same referent is easier than doing the same with two phonologically dissimilar labels (non-cognates, like *door* and *puerta*). If cognates are easier to acquire than non-cognates, bilinguals learning two languages that share a high proportion of cognates should benefit more often from this facilitation effect than bilinguals learning two languages with a lower proportion of cognates, and should therefore show larger vocabulary sizes. @floccia2018introduction provided evidence for such a facilitation effect of typological distance and vocabulary size. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh.

The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Phonological overlap was measured by computing the Levenshtein distance between each cross-language pair of phonetic transcriptions. The Levenshtein distance is a metric that computes the distance between two strings of characters/phonemes by counting the smallest number of insertions, deletions and substitutions one of the strings has to undergo to become identical to the other [@levenshtein1966binary]. For instance, the Levenshtein distance for the translation equivalent *cat*-*kat* (/\textipa{"kæt}-\textipa{"kAt}/) in English and Dutch would be one: a single edition would be needed to make both phonological translation identical (replacing /\textipa{æ}-\textipa{A}/). The same translation equivalent in English and Greek (*cat*-*$\gamma\acute{\alpha}\tau\alpha$*, /\textipa{"kæt}-\textipa{"Ga.ta}/) would score three, as it requires at least three editions: substituting /\textipa{k}/ for /\textipa{G}/, substituting /\textipa{\ae}/ for /\textipa{a}/ and adding another /\textipa{a}/. Levenshtein distance scores are divided by the length of the longest string to normalise the similarity scores between 0 and 1. Finally, the result is subtracted from one, to get the complementary, and therefore a measure of phonological similarity, instead of dissimilarity. Following the previous examples: $1/4 = 0.25$ for /\textipa{"kæt}-\textipa{"kAt}/, and $3/4 = 0.75$ for /\textipa{"kæt}-\textipa{"Ga.ta}/. Finally, this variable was included as a predictor of participants' vocabulary sizes. Among other findings, Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (0.2214 mean phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (0.0197 mean phonological overlap) were able to produce in Mandarin. More recently, @blom2020crosslanguage and @bosma2019longitudinal reported similar results, providing converging evidence of for facilitatory effect of language distance on vocabulary size estimates.

Floccia et al. pointed to *parallel activation* as the underpinning mechanism behind their results. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during speech production or comprehension, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access is not restricted to production or the auditory modality, but rather extends to comprehension and translation, and to the visual modality (reading and signing) [@spivey1999cross; @hoshino2008cognate; @schwartz2007reading; @schwartz2006bilingual; @morford2011when; @vanhell2008sentence; @gimeno-martinez2021crosslinguistic; see @kroll2013multilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*), importantly such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. These results showed that bilinguals' Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures.

Bilingual toddlers and children---still in the process to developing a mature lexicon---also show evidence of parallel activation during comprehension [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical]. Whether---and if so, how---the language non-selectivity shapes the early bilingual lexicon is still unclear. Previous work has pointed to an interactive account of lexical acquisition, in which toddlers' lexical representations in one language influence the acquisition of words in the other language. This cross-language facilitation is, in principle, possible since early ages: the bilingual infant lexicon consists in a substantial proportion of  translation equivalents: the presence of a word in the lexicon is a good predictor of its translation equivalent also being present [@bilson2015semantic; @tsui2022are; @pearson1995crosslanguage; @pearson1995crosslanguage]. In fact, translation equivalents may even be overrepresented in the early bilingual lexicon. @bilson2015semantic found that bilingual toddlers produced a higher proportion of translation equivalents than it would be expected if they learned two independent lexica simultaneously, which suggests that bilingual children preferentially acquired words for which they already knew a label in the other language. In the same line, @bosch2014first reported a similar overrepresentation of translation equivalents in Catalan-Spanish bilinguals: a high proportion of acquired Catalan words held an also acquired translation in Spanish. Finally, a recent investigation by @tsui2022are presented a quantitative model of bilingual vocabulary acquisition which showed a consistent preferential acquisition of translation equivalents by English-French bilinguals, with such preference being stronger at early stages of vocabulary growth. Overall, these findings suggest that bilingual children benefit from a semantic facilitation effect driven by the acquisition of translation equivalents.

If translation equivalents are pervasive in the bilingual lexicon since early ages, it is possible that cognateness increases the amount of cross-language activation, boosting early word acquisition. This could ultimately lead to children who are learning languages with a larger proportion of cognates (e.g., English and Dutch) showing larger vocabulary sizes, as found by Floccia et al. There is evidence of phonological and semantic similarity facilitating word acquisition in monolinguals: infants are more likely to acquired new words that are semantically or phonologically more connected to other words, than words with poorer connectivity in either of those dimensions, such as words belonging to sparse phonological neighbourhoods [@hills2009longitudinal; @fourtassi2020growth; @laing2022phonological; @jones2019children; @storkel2004children]. If phonological similarity plays a cross-language facilitation role during bilingual word acquisition, cognate translation equivalents---which share both semantic and phonological similarity---should be acquired, on average, earlier that non-cognate translation equivalents---which share semantic similarity (e.g., are taxonomically related) but not phonological similarity.

Evidence supporting an earlier age-of-acquisition for cognates is scarce. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months, and found that cognates represented a larger proportion of participant's vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. The low sample size in these two studies makes it challenging to draw strong conclusions about the effect of cognateness on vocabulary growth. Floccia et al.'s estimates are statistically more reliable given their remarkably larger sample size (372 participants). Although their results suggest a cognateness effect on vocabulary size, their study was not aimed at testing the effect of cognateness on age-of-acquisition at the *word-level*. The response variable in this study was the proportion of words each participant understood and/or produced (i.e., vocabulary size), from the list of lexical items in the vocabulary checklists. By aggregating the responses from all words into a single datum per child, it was no longer possible to test the effect of cognateness on word acquisition at the word-form level. Also, all participants were aged 24 months, meaning that even if the unaggregated responses to individual items were included as the response variable, age-sensitive claims about the effect of cognateness on word acquisition should be taken with caution [a limitation also present in the @bosch2014first study].

More recently, @mitchell2022cognates addressed this issue providing a more fine grained measure of a  cognateness facilitation effect in vocabulary acquisition. Using a longitudinal sample of 47 16-to-30 month-old French-English bilinguals living in Canada, the authors collected data on expressive vocabulary data in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of translation equivalents that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). These findings shed some light on the ongoing exploration of why (if at all) bilinguals' vocabulary size seems to grow faster when both languages are more similar at the lexical level. Word-forms sharing more phonemes with their translation equivalents (i.e., cognates) seem to be acquired faster. Parallel activation remains a plausible explanation for this effect. When a child is exposed to a word-form, its (phonologically similar) translation equivalent is activated, and this cross-language activation increases the chances of acquisition. The precise underpinnings of this effect are still unclear.

In the present investigation we approach the cognate facilitation on word learning by extending the conceptual framework of accumulator models of language acquisition to the bilingual case. Accumulator models formalise the widespread assumption that a child's vocabulary growth rate is a function of mainly---but not exclusively---the number of learning instances that a child has accrued with a particular word [see @kachergis2022standard for review]. A learning instance of a word consists, for example, in hearing such word-form (*flower*) in the presence of its referent (a flower). We devise word acquisition as a continuous process in which a child accumulates experience with a given word-form and its referential context until the number of cumulative learning instances meets some theoretical threshold. After this, we consider the lexical representation of the word as *consolidated*, which leads to a non-linear increase in the probability of caregivers reporting such word as acquired in a vocabulary checklist like the one in the present study. As infants receive linguistic input, they also accumulate opportunities to learn words until a threshold is crossed, an the word can be considered to be acquired. The rate at which monolingual children encounter learning instances with some word-form is strongly determined by age: as time passes children are exposed to a larger amount of linguistic input, and therefore accumulate more learning experiences with all words.

Not all words occur in the linguistic input with identical frequency, so children accumulate experience with some word-forms at faster rates than with other word-forms. This results in frequent words being acquired earlier in age than less frequent words [e.g., @ambridge2015ubiquity; @goodman2008does; @braginsky2019consistency; @swingley2018quantitative; @hansen2017what; @verhagen2022determinants]. For this reason, lexical frequency plays a central role in most models of lexical growth, especially in the case of accumulator models [@hidaka2013computational; @mollica2017how]. Following this account, words with high lexical frequency like 'flower' [`r format(floor(childes[childes$token=="flower",]$freq_million), big.mark = ",")` times per million words in the CHILDES corpus, @macwhinney2000childes] are predicted to be acquired earlier in age than words with low lexical frequency such as 'taxi' (`r format(floor(childes[childes$token=="taxi",]$freq_million), big.mark = ",")` times per million words).

We argue that previous accounts on vocabulary growth based on accumulator models are a convenient framework to study bilingual vocabulary growth. Bilinguals share fundamentally similar learning mechanisms with their monolinguals peers. Developmental divergences between both groups can be understood as the result of the the negotiation between such mechanisms and the particular demands of a dual language input [e.g., @werker2005primir; @curtin2011bilingual]. One of such divergences is the fact that bilinguals' frequency of exposure to a word-form is mediated by their degree of exposure to their languages. Assuming that bilinguals and monolinguals are exposed to an equivalent amount of linguistic input, a Spanish-English bilingual who is exposed to Spanish 80% of the time and to English 20% of the time would need more time to accumulate the same experience with 'flower' and 'taxi' than an English monolingual (who listens 100% of the time to English) would need.

In the present study, we incorporate this feature of the bilingual linguistic input to an account of the cognateness effect on word acquisition. We suggest that when a bilingual child is exposed to a given word, they not only activate its corresponding lexical representation, but also that of its translation equivalent. This co-activation is proportional to the form similarity between both word-forms. When this parallel activation is in place, each exposure to one of the word-forms may count as a learning instance for both members of the translation equivalent, in a degree determined by their cognateness. Cognate translation equivalents, for which stronger co-activation is expected, accumulate learning instances faster than non-cognates, for which weaker or no co-activation is expected. Finally,  when the child has accumulator some threshold number of learning instances with any of these word-forms, its lexical representation reaches some degree of consolidation that leads to a non-linear change in the child's observed learning outputs, like the probability of such word being reported as "acquired" by caregivers, or by stronger naming effects in experimental tasks using the Visual World Paradigm [e.g., @swingley2002recognition]. Because , lexical representations of cognate words accumulate learning instances faster than those of non-cognate words, the former would consolidate earlier in age.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], whose word-forms share three phonemes (0.75 Levenshtein similarity). When the child is exposed to /\textipa{"gat}/, they might co-activate /\textipa{"ga.to}/ in parallel, due to their phonological similarity. If this exposure to the word form occurs in a convenient referential context (e.g., in the presence of a cat), this might count as a learning instance for both active word-forms. Consider now the case of the non-cognate translation equivalent /\textipa{"gos}-\textipa{"pe.ro}/. Because of their low phonological similarity (0% Levenshtein similarity), the amount of co-activation of both words might be low, which might leads to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for  /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ are predicted to benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ are not. If the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. If parallel activation is in place, the former will activate the latter more often than vice versa. Ultimately, /\textipa{"ga.to}/ should benefit more strongly from its cognate status than /\textipa{"gat}/. For instance, if the previous child is exposed 80% of the time to Catalan, and 20% of the time to Spanish, words in Spanish should, on average, show an even stronger cognateness facilitation effect than words in Catalan.

@fig-hypotheses provides a graphic, theoretical illustration of this account. We assume the case of a Catalan-Spanish bilingual child who is exposed 60% of the time to Catalan, and 40% of the time to Spanish. We also assume that the child is exposed to either word around 50 times per month, following a $\text{Poisson}(\lambda = 50)$ distribution. We illustrate the cumulative sum of learning instances that a child accumulates across ages for the Catalan word-form (/\textipa{"gat}/ or /\textipa{"gos}/) and its Spanish translation (/\textipa{"pe.ro}/ or /\textipa{"ga.to}/). We assume an arbitrary, theoretical threshold for the learning instances at which the child is considered to have acquired a word (solid grey line). Points and triangles show the age-of-acquisition at which the amount of learning instances reaches the threshold. We show the predicted age-of-acquisition under two hypothesis: (1) a cognateness facilitation effect through parallel activation, in which both word-forms co-activate each other accumulating learning instances faster, and (2) no cognateness facilitation, for which both word-forms accumulate learning instances independently.

```{r fig-hypotheses}
#| label: fig-hypotheses
#| fig-cap: "Schematic representation of the hypothesised cognate facilitation effect on word acquisition."
#| fig-width: 8
#| fig-height: 7
#| out-width: 100%
#| out-height: 100%
#| echo: false
#| fig-align: center
threshold <- 300

# no parallel activation
eli_df <- generate_eli(
    threshold = threshold,
    age = seq(1, 50, length.out = 100),
    freq_month = 50,
    freq_beta = 1,
    l1_doe = 0.60,
    parallel_beta = 0,
    conditions = c(
        glue::glue("**Cognate**: /<i style='color:{clrs[1]}'>ˈgat</i>-<i style='color:{clrs[4]}'>ˈga.to</i>/ [cat]"),
        glue::glue("**Non-cognate**: /<i style='color:{clrs[1]}'>ˈgos</i>-<i style='color:{clrs[4]}'>ˈpe.ro</i>/ [dog]")
    )
) |>
    filter(hypothesis=="H0") |> 
    mutate(language = ifelse(language=="L1", "Catalan form", "Spanish form"))

aoa_df <- generate_aoa(eli_df, threshold = threshold) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H0")
# 
# img <- c(cat = "images/diagram_cat.png",
#          dog = "images/diagram_dog.png") |>
#     map(magick::image_read) |>
#     map(\(x) magick::image_ggplot(x, interpolate = FALSE))

plot_h0 <- eli_df |>
    ggplot(aes(age, eli,
               colour = language)) +
    facet_grid(~ te) +
    geom_segment(data = aoa_df,
                 aes(x = aoa,
                     xend = aoa,
                     y = 0,
                     yend = threshold),
                 linewidth = 3/4) +
    geom_hline(yintercept = aoa_df$threshold,
               linewidth = 3/4,
               colour = "grey80") +
    annotate(geom = "label",
             label = glue::glue("Threshold: {threshold}"),
             label.r = unit(0, "lines"),
             label.size = 0,
             fill = "grey80",
             x = 0,
             y = threshold,
             size = 2,
             hjust = 0) +
    geom_line(linewidth = 1) +
    geom_label(data = aoa_df,
               aes(x = aoa,
                   y = threshold,
                   label = round(aoa, 2),
                   fill = language),
               size = 3,
               label.r = unit(0, "lines"),
               label.size = 0,
               colour = "white",
               na.rm = TRUE,
               show.legend = FALSE
    ) +
    labs(x = "Age (months)",
         y = "Cumulative\nlearning instances",
         colour = "",
         shape = "Hypothesis",
         fill = "",
         title = "H0: no parallel activation",
         linetype = "Hypothesis") +
    scale_colour_manual(values = clrs[c(1, 4)]) +
    theme(legend.position = "bottom",
          legend.key.width = unit(1.5, "cm"),
          legend.margin = margin(c(0, 0, 0, 0)),
          legend.justification = c(0, 1),
          legend.title = element_text(size = 9),
          legend.text= element_text(size = 8))

# with parallel activation
eli_df <- generate_eli(
    threshold = threshold,
    age = seq(0, 50, by = 1),
    l1_doe = 0.60,
    freq_month = 50,
    freq_beta = 1,
    conditions = c(
        glue::glue("**Cognate**: /<i style='color:{clrs[1]}'>ˈgat</i>-<i style='color:{clrs[4]}'>ˈga.to</i>/ [cat]"),
        glue::glue("**Non-cognate**: /<i style='color:{clrs[1]}'>ˈgos</i>-<i style='color:{clrs[4]}'>ˈpe.ro</i>/ [dog]")
    )
) |>
    filter(hypothesis=="H1") |> 
    mutate(language = ifelse(language=="L1", "Catalan form", "Spanish form"))

aoa_df <- generate_aoa(eli_df, threshold = threshold) |>
    mutate(across(aoa, lst(min, max)), .by = c(te, language)) |>
    filter(hypothesis=="H1")

>>>>

abstract: |
    Recent work on bilingual language acquisition points to cognate words being acquired earlier in age than non-cognates. Cross-language activation has been proposed as a candidate mechanism behind this facilitation effect, but the specific dynamics of lexical Building on accumulator models of language acquisition, we present an account of word acquisition in which the co-activation of cognate translation equivalents increases the rate at which of bilingual children encounter learning instances for words in both languages. We tested these predictions against vocabulary data from 366 Catalan-Spanish bilinguals toddlers. We used Bayesian Explanatory Item Response Models to estimate participants’ probability of acquisition of 604 words, conditional to the cognate status and lexical frequency of the word-form, and the age and degree of exposure to each language of the participant. We found a cognateness facilitation effect only in low-frequency words or words from the language of lower exposure. Overall, our findings support an interactive account of bilingual vocabulary acquisition in which the lexical representations in one language interact with the acquisition of words in the other language.

@@

The foundations of word learning are in place early in age. From six months of age, infants start directing their gaze to some objects when hearing their labels, according to experimental data [@tincoff1999beginnings; @bergelson2012months; @bergelson2015early]. Shortly after, caregivers start reporting some words as acquired by their infant in vocabulary checklists [e.g., @fenson2007macarthurbates; @samuelson2021precision]. Most research on early word acquisition relies extensively on data from monolingual children and is oblivious to the fact that a substantial proportion of the world population acquires more than one language from early ages [@grosjean2021extent]. Previous work on bilingual vocabulary acquisition pointed to bilingual toddlers knowing, on average, less words in each of their languages than their monolinguals peers, and to both groups knowing a similar amount of words, if not more words, when the bilinguals’ two languages are pooled together. @hoff2012dual found that English-Spanish bilingual toddlers in South Florida (United States) knew less words in English than monolinguals did. Both groups knew a similar total amount of words when both English and Spanish vocabularies were counted together. Other studies have provided converging evidence that bilinguals know a similar or even larger number of words than monolinguals when the two languages are aggregated [@oller2002language; @pearson1994patterns; @pearson1993lexical; @patterson2004comparing; @patterson2004bilingual; @smithson2014bilingualism; @petitto2001bilingual; @gonzalez-barrero2020bilingual]. A more detailed analysis of the words in bilinguals’ lexicon show some interesting patterns.

One important observation of studies on bilinguals’ early vocabulary acquisition is that cognate words are easier to acquire than non-cognate words. Cognate words are translation equivalents that are phonologically similar (or share some type of form-similarity). For instance, the Spanish translation equivalent of cat is gato, a cognate word; the translation equivalent of dog is perro, a non-cognate word. The differences in the percentages of cognate and non-cognate words between two languages is related to historic reasons: languages typologically close (like Dutch and English or Italian and Spanish) share more cognates than languages typologically distant (like English and Chinese, or Urdu and Spanish). The conclusion that cognate words are easier to learn is based on two types of evidence: studies investigating vocabulary sizes in children learning language pairs with different percentages of cognates (that is, differing in their typological distance) and studies comparing the number of cognate and non-cognate words children know in a specific language pair.

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi/Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish and Welsh. The authors calculated the average phonological overlap between the words in each of these additional languages and their translation equivalents in English, which was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of language distance on vocabulary size.

A second set of studies concluded that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of participant’s vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual produced cognates earlier than non-cognates, on average. @mitchell2022cognates addressed this issue in a larger sample longitudinal study. These authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages, even when both lists were matched by semantic category (furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research provide a strong support to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual environments. Evidence supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @vanhell2008sentence; @hoshino2008cognate; @morford2011when; @schwartz2007reading; @spivey1999cross; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (árbol-arbre, translations of tree), whereas in the other half of the trial labels were non-cognates (mesa-taula, translations of table), obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. Evidence of parallel activation has been reported in bilingual toddlers and children [@vonholzen2012language; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @schroter2016orthographic; @poulin-dubois2013lexical].

Although there is converging evidence on the role of parallel activation, the previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words, and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, only if such translation equivalent is co-activated. The probability of this co-activation is proportional to the phonological similarity between the two translation equivalents, and given that cognates share higher phonological similarity than non-cognates, the former may be co-activated more frequently than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across both languages: words from the lower-exposure language would receive additional learning instances through parallel activation more often than words from the higher-exposure language. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}-\textipa{"ga.to}/ [cat], that  are phonologically very similar. When the child is exposed to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. Therefore, this exposure will count as a learning instance for both co-activated forms. However, for the case of the non-cognate translation equivalent /\textipa{"gos}/, the low phonological similarity will result in a weak activation of /\textipa{"pe.ro}/ resulting in such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}-\textipa{"ga.to}/ will benefit from parallel activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children using an online vocabulary checklist designed for the present study. We adopted a Bayesian explanatory item response theory approach (IRT, see @kachergis2022standard, for a similar approach) to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition. We included participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency). We predicted an interaction between cognate status and word exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure words cognate words.
